---
title: "Chapter 9: Functionals"  
author: "Erika Duan"
date: "`r Sys.Date()`"
output: 
  github_document:
    toc: true
---  

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, results = 'hide')
```

```{r, message = FALSE, warning = FALSE}
# load R libraries -------------------------------------------------------------   
if (!require(pacman)) install.packages("pacman")
p_load(tidyverse)  
```


# Chapter goals   

Understand how to adopt best functional programming practices including:  

+ Understand the definition of a pure function.  
+ Learn how to decompose lengthy code into isolated and pure functions that operate independently.  
+ Replace for loops with generalised functions that take another function as an argument. 


# Fundamental concepts of functional programming       

Functional programming languages have:  

+ First-class functions - full versatility to handle functions as if they are just another data structure i.e. they can be assigned, stored in lists, passed as arguments to other functions or created inside other functions.     
+ Pure functions - reproducible and consistent outputs are generated from the same inputs, with no side-effects (i.e. global environment modifications) triggered.  

**Note:** Exception to pure functions include `runif()`, `sample()` and `Sys.time()` due to differing output generation, and `write.csv()` and `<-` due to global environment modification, although you can argue that the intention of these functions is always consistent.    

